Multiset-Trie Data Structure

### 2023-03-15 tex/algorithms-2122309-proof.tex
a few updates...
updates for the %MDPI comments. numbers remain to be corrected.

% Abstract (Do not insert blank lines, i.e., \\) 
\abstract{This paper proposes a new data structure, \highlighting{\emph{multiset-trie}}, %MDPI: Please confirm if the italics is unnecessary and can be removed. Please check all italic.
 that is designed for storing and efficiently processing a set of multisets. Moreover, multiset-trie can operate on a set of sets without efficiency loss. 
The multiset-trie structure is a search tree with properties similar to those of a trie. It implements all standard search tree operations together with the multiset containment operations, such as sub-multiset and super-multiset. Suppose that we have a set of multisets $S$ and a multiset $X$. The multiset containment operations retrieve multisets from $S$ that are either sub-multisets or super-multisets of $X$.  
We present the mathematical analysis of a multiset-trie that gives the time complexity of the algorithms and the space complexity of the data structure. Further, the empirical analysis of the data structure is implemented in a series of experiments. The experiments illuminate the time complexity space of the multiset containment operations.
%Finally, the comparison of the multiset-trie with the inverted index shows that the multiset-trie outperforms the inverted index by up to 4 orders of magnitude.
%%% more rounded...!
}

@@

%
Let $\mathcal{M}$ be a multiset-trie and let $M$ be a set of multisets that are 
inserted into the multiset-trie $\mathcal{M}.$ We define a type \emph{Multiset} in 
order to use it as a representation of a multiset. The type \emph{Multiset} is 
an array $m$ of constant length $\sigma,$ where the $i$-th cell represents the element 
$\phi^{-1}(i)$ from $\Sigma$ with multiplicity $m[i].$ From now on, we 
agree that the first cell of an array has index 1. Let us give an example of a 
\emph{Multiset} instance with $\sigma = 2:$
%
\begin{center}
\begin{tabular}{ccc}
Multiset & & Instance of type Multiset \\
$\{ 1,1,2 \}$ & $\cong $ & 
\begin{tabular}{|c|c|}
\hline 
2 & 1 \\
\hline 
\multicolumn{1}{c}{\tiny 1} & \multicolumn{1}{c}{\tiny 2} \\
\end{tabular}
\end{tabular}
\end{center}
%
\hl{The} %MDPI: Whether the indentation of the paragraph after the formula is necessary, please confirm.
 operations supported by the multiset-trie data structure are as
follows. 
%
\begin{enumerate}
\item \textsc{insert}($\mathcal{M}$, $m$): inserts a multiset $m$ into 
$\mathcal{M}$ if $m\not\in M;$
%
\item \textsc{search}($\mathcal{M}$, $m$): returns true if a multiset $m\in M$ 
for a given $\mathcal{M},$ and returns false otherwise;
%
\item \textsc{delete}($\mathcal{M}$, $m$): returns true if a multiset $m$ was 
successfully deleted from $\mathcal{M},$ and returns false otherwise (in case 
$m\not\in M$);
%
\item \textsc{submsetExistence}($\mathcal{M}$, $m$, $dev$): returns true if 
there exists a $x\in M$ for a given $\mathcal{M}$ such that $x\subseteq m$ 
and $| x[i] - m[i] |\leq dev$ for $1\leq i\leq \sigma$, and returns false otherwise; 
%
\item \textsc{supermsetExistence}($\mathcal{M}$, $m$, $dev$): returns true if 
there exists a $x\in M$ for a given $\mathcal{M}$ such that $x\supseteq m$ 
and $| x[i] - m[i] |\leq dev$ for $1\leq i\leq \sigma$, and returns false otherwise; 
%
\item \textsc{getAllSubmsets}($\mathcal{M}$, $m$, $dev$): returns the set of multisets 
$\{ x \in M : x\subseteq m \wedge |x[i]-m[i]|\leq dev \}$ for a given 
$\mathcal{M},$ where $1\leq i\leq \sigma;$
%
\item \textsc{getAllSupermsets}($\mathcal{M}$, $m$, $dev$): returns the set of multisets 
$\{ x\in M : x\supseteq m \wedge |x[i]-m[i]|\leq dev \}$ for a given $\mathcal{M},$
where $1\leq i\leq \sigma.$
%
\end{enumerate}

@@

\includegraphics[width=.64\textwidth, keepaspectratio]{ratio-exp-nodes-and-msets-on-prob1.pdf}
\caption{The ratio of ``data overhead'' derived from Figure \ref{f:exp-nodes}, as the density $p$ of our data structure increases.
Unless we are in a very sparse setting, the trade-off in terms of space is small.}
\label{f:ratio-exp-msets}
\end{figure}

@@

%
This section contains the results of experiments that were performed on the multiset-trie 
data structure. In particular, we tested the functions \textsc{submsetExistence}, 
\textsc{supermsetExistence}, \textsc{getAllSubmsets}, and \textsc{getAllSupermsets}.

@@

\subsection{\hl{Data} %MDPI: Section headings should be numbered sequentially, e.g., Section 2.1, Section 2.2.1. Please confirm this revision.
 Generation}
We denote by \emph{input data} the data that are used to fill the structure prior 
to testing and by \emph{test data} the set of queries that are used to test the 
performance of the functions.

@@

\subsection{Experiment 1} \label{s:exp1}
This experiment shows the performance of multiset-trie being used for storing 
and retrieving \emph{sets} instead of \emph{multisets}. We restrict multiset-trie in order 
to obtain a closer comparison with the \emph{set-trie} data structure~\cite{savnik2013index}.
In this case, we set the maximal node degree $n$ to be $2$ and $\sigma$ to be 25. 
The mapping $\phi$ does not have an influence in this particular experiment
because the input data are generated artificially with uniform distribution. On 
average, the results will be the same for any $\phi,$ since all the multisets are 
equally likely to appear in $M.$ The parameter $|M|$ varies from \hl{10,000} %MDPI: We added commas to separate thousands for numbers with five or more digits. Please confirm.
 sets up 
to \hl{320,000} sets. According to the parameters $n$ and $\sigma,$ the cardinality of 
$N$ is $33,554,432\approx \num{3.36e+7}.$ Thus, the calculated density of the 
multiset-trie with respect to $|M|$ varies from~$\num{0.3e-3}$ to~$\num{9.5e-3}.$

@@

\begin{figure}[H]
	\subcaptionbox{\centering  \label{fig:e1m1}}{\includegraphics[width=.485\textwidth]{exp1-m1.pdf}}
	\subcaptionbox{\centering \label{fig:e1m2}}{\includegraphics[width=.485\textwidth]{exp1-m2.pdf}}
	\caption{\hl{Existence} %MDPI: Please use commas to separate thousands for numbers with five or more digits (not four digits) in the picture, e.g., "10000" should be "10,000".
 functions of Experiment 1. (\textbf{a}) \hl{submsetExistence;} %MDPI: We moved the subfigure explanations into the figure caption. Please confirm. The following are same.
 (\textbf{b}) \hl{supermsetExistence}.  \label{fig:4}}
\end{figure}

@@

\subsection{Experiment 2} \label{s:exp2}
In Experiment~\hyperref[s:exp2]{2}, we demonstrate the performance of 
the unrestricted multiset-trie allowing \emph{multisets} to be inserted into the data structure. 
We set $n$ to be 6 and retain $\sigma = 25$, as in Experiment~\hyperref[s:exp1]{1}. 
The mapping $\phi$ does not have an influence on the results, since the input 
data are generated artificially with uniform distribution. The cardinality of $M$ 
varies from \hl{40,000} to \hl{640,000} multisets. Thus, the calculated density $d$ varies 
from $\num{1.4e-15}$ to $\num{2.25e-14}.$ The density is much smaller than 
in Experiment~\hyperref[s:exp1]{1}, because now we allow multisets to be stored 
in the data structure, and according to the parameters $n$ and $\sigma$, the 
cardinality of $N$ is $6^{25} = \num{2.84e19}.$

@@

\subcaptionbox{\centering}{\includegraphics[width=.485\textwidth]{exp4-m1.pdf}}
\subcaptionbox{\centering 
\label{fig:e3m2}}{\includegraphics[width=.485\textwidth]{exp4-m2.pdf}}
\caption{Existence functions in Experiment 3. (\textbf{a}) \hl{submsetExistence}; (\textbf{b}) supermsetExistence.\label{fig:e3m1}}
\end{figure}

@@

The data are taken from the English dictionary, which contains \hl{235,883} different words. 
These words are mapped to multisets of integers according to the $\phi.$ In 
particular, we are interested in cases where $\phi(\Sigma)$ enumerates 
letters by their relative frequency in the English language. We say that $\phi(\Sigma)$ 
maps letters in \emph{ascending order} if the most frequent letter is mapped to 
number $\sigma.$ Conversely, in \emph{descending order}, this letter is mapped to 
the number $1.$ The size of the alphabet $\sigma$ is set to the size of the English 
alphabet: 26. The degree of a node $n$ is set to 10. On average, the multiplicity 
of letters is, of course, less than 10. We choose such a large node degree allowing 
the multiplicity to be up to 10 because the dictionary contains such words.

@@

The experiment uses the input data for the construction of the given
data structure and the test data for the execution of the operations
on the given data structure. The input data comprise a set of
randomly generated multisets used for the construction of a data
structure. The test data include the set of multisets together with
the operations that are evaluated. The input and test data were
generated with respect to parameters $\sigma$ and $n$, as presented in
\hl{x} %MDPI: Please confirm if this can be removed.
 Table~\ref{t:benchmark}.

\begin{table}[H]
\caption{\hl{Configuration} %MDPI: We removed the vertical lines. Please confirm.
 of $\sigma$ and $n$ in benchmark.}
\label{t:benchmark}
\newcolumntype{C}{>{\centering\arraybackslash}X}
\begin{tabularx}{\textwidth}{CC}
\toprule
\boldmath$\sigma$ & \boldmath$n$ \\
\midrule
5		& 1\\
%\hline
30	& 1 \\
%\hline
5		& 3 \\
%\hline
15	& 3 \\
%\hline
30	& 3 \\
%\hline
10	& 10 \\
\bottomrule
\end{tabularx}
\end{table}

>>>>

% Abstract (Do not insert blank lines, i.e., \\) 
\abstract{This paper proposes a new data structure, multiset-trie, %MDPI: Please confirm if the italics is unnecessary and can be removed. Please check all italic. %RES Italics removed.
 that is designed for storing and efficiently processing a set of multisets. Moreover, multiset-trie can operate on a set of sets without efficiency loss. 
The multiset-trie structure is a search tree with properties similar to those of a trie. It implements all standard search tree operations together with the multiset containment operations, such as sub-multiset and super-multiset. Suppose that we have a set of multisets $S$ and a multiset $X$. The multiset containment operations retrieve multisets from $S$ that are either sub-multisets or super-multisets of $X$.  
We present the mathematical analysis of a multiset-trie that gives the time complexity of the algorithms and the space complexity of the data structure. Further, the empirical analysis of the data structure is implemented in a series of experiments. The experiments illuminate the time complexity space of the multiset containment operations.
%Finally, the comparison of the multiset-trie with the inverted index shows that the multiset-trie outperforms the inverted index by up to 4 orders of magnitude.
%%% more rounded...!
}

@@

%
Let $\mathcal{M}$ be a multiset-trie and let $M$ be a set of multisets that are 
inserted into the multiset-trie $\mathcal{M}.$ We define a type \emph{Multiset} in 
order to use it as a representation of a multiset. The type \emph{Multiset} is 
an array $m$ of constant length $\sigma,$ where the $i$-th cell represents the element 
$\phi^{-1}(i)$ from $\Sigma$ with multiplicity $m[i].$ From now on, we 
agree that the first cell of an array has index 1. Let us give an example of a 
\emph{Multiset} instance with $\sigma = 2:$
%
\begin{center}
\begin{tabular}{ccc}
Multiset & & Instance of type Multiset \\
$\{ 1,1,2 \}$ & $\cong $ & 
\begin{tabular}{|c|c|}
\hline 
2 & 1 \\
\hline 
\multicolumn{1}{c}{\tiny 1} & \multicolumn{1}{c}{\tiny 2} \\
\end{tabular}
\end{tabular}
\end{center}

The %MDPI: Whether the indentation of the paragraph after the formula is necessary, please confirm. %RES Indentation added.
 operations supported by the multiset-trie data structure are as
follows. 
%
\begin{enumerate}
\item \textsc{insert}($\mathcal{M}$, $m$): inserts a multiset $m$ into 
$\mathcal{M}$ if $m\not\in M;$
%
\item \textsc{search}($\mathcal{M}$, $m$): returns true if a multiset $m\in M$ 
for a given $\mathcal{M},$ and returns false otherwise;
%
\item \textsc{delete}($\mathcal{M}$, $m$): returns true if a multiset $m$ was 
successfully deleted from $\mathcal{M},$ and returns false otherwise (in case 
$m\not\in M$);
%
\item \textsc{submsetExistence}($\mathcal{M}$, $m$, $dev$): returns true if 
there exists a $x\in M$ for a given $\mathcal{M}$ such that $x\subseteq m$ 
and $| x[i] - m[i] |\leq dev$ for $1\leq i\leq \sigma$, and returns false otherwise; 
%
\item \textsc{supermsetExistence}($\mathcal{M}$, $m$, $dev$): returns true if 
there exists a $x\in M$ for a given $\mathcal{M}$ such that $x\supseteq m$ 
and $| x[i] - m[i] |\leq dev$ for $1\leq i\leq \sigma$, and returns false otherwise; 
%
\item \textsc{getAllSubmsets}($\mathcal{M}$, $m$, $dev$): returns the set of multisets 
$\{ x \in M : x\subseteq m \wedge |x[i]-m[i]|\leq dev \}$ for a given 
$\mathcal{M},$ where $1\leq i\leq \sigma;$
%
\item \textsc{getAllSupermsets}($\mathcal{M}$, $m$, $dev$): returns the set of multisets 
$\{ x\in M : x\supseteq m \wedge |x[i]-m[i]|\leq dev \}$ for a given $\mathcal{M},$
where $1\leq i\leq \sigma.$
%
\end{enumerate}

@@

\includegraphics[width=.64\textwidth, keepaspectratio]{ratio-exp-nodes-and-msets-on-prob.pdf}
\caption{The ratio of ``data overhead'' derived from Figure \ref{f:exp-nodes}, as the density $p$ of our data structure increases.
Unless we are in a very sparse setting, the trade-off in terms of space is small.}
\label{f:ratio-exp-msets}
\end{figure}

@@

%
This section contains the results of experiments that were performed on the multiset-trie 
data structure. In particular, we tested the functions \textsc{submsetExistence}, 
\textsc{supermsetExistence}, \textsc{getAllSubmsets}, and \textsc{getAllSupermsets}. 
We refer to the former two operations as the \emph{existence operations}, and the
later two as the \emph{exhaustive operations}. %RES A sentence added to introduce the terms existence and exhaustive operations used later.

@@

\subsection{Data%MDPI: Section headings should be numbered sequentially, e.g., Section 2.1, Section 2.2.1. Please confirm this revision. %RES It is Ok as it is.
 Generation}
We denote by \emph{input data} the data that are used to fill the structure prior 
to testing and by \emph{test data} the set of queries that are used to test the 
performance of the functions.

@@

\subsection{Experiment 1} \label{s:exp1}
This experiment shows the performance of multiset-trie being used for storing 
and retrieving \emph{sets} instead of \emph{multisets}. We restrict multiset-trie in order 
to obtain a closer comparison with the \emph{set-trie} data structure~\cite{savnik2013index}.
In this case, we set the maximal node degree $n$ to be $2$ and $\sigma$ to be 25. 
The mapping $\phi$ does not have an influence in this particular experiment
because the input data are generated artificially with uniform distribution. On 
average, the results will be the same for any $\phi,$ since all the multisets are 
equally likely to appear in $M.$ The parameter $|M|$ varies from 10,000 %MDPI: We added commas to separate thousands for numbers with five or more digits. Please confirm. %RES We confirm.
 sets up 
to 320,000 sets. According to the parameters $n$ and $\sigma,$ the cardinality of 
$N$ is $33,554,432\approx \num{3.36e+7}.$ Thus, the calculated density of the 
multiset-trie with respect to $|M|$ varies from~$\num{0.3e-3}$ to~$\num{9.5e-3}.$

@@

\begin{figure}[H]
	\subcaptionbox{\centering  \label{fig:e1m1}}{\includegraphics[width=.485\textwidth]{exp1-m1.pdf}}
	\subcaptionbox{\centering \label{fig:e1m2}}{\includegraphics[width=.485\textwidth]{exp1-m2.pdf}}
	\caption{\hl{Existence} %MDPI: Please use commas to separate thousands for numbers with five or more digits (not four digits) in the picture, e.g., "10000" should be "10,000".
 functions of Experiment 1. (\textbf{a}) submsetExistence; %MDPI: We moved the subfigure explanations into the figure caption. Please confirm. The following are same. %RES We confirm.
 (\textbf{b}) supermsetExistence.  \label{fig:4}}
\end{figure}

@@

\subsection{Experiment 2} \label{s:exp2}
In Experiment~\hyperref[s:exp2]{2}, we demonstrate the performance of 
the unrestricted multiset-trie allowing \emph{multisets} to be inserted into the data structure. 
We set $n$ to be 6 and retain $\sigma = 25$, as in Experiment~\hyperref[s:exp1]{1}. 
The mapping $\phi$ does not have an influence on the results, since the input 
data are generated artificially with uniform distribution. The cardinality of $M$ 
varies from 40,000 to 640,000 multisets. Thus, the calculated density $d$ varies 
from $\num{1.4e-15}$ to $\num{2.25e-14}.$ The density is much smaller than 
in Experiment~\hyperref[s:exp1]{1}, because now we allow multisets to be stored 
in the data structure, and according to the parameters $n$ and $\sigma$, the 
cardinality of $N$ is $6^{25} = \num{2.84e19}.$

@@

\subcaptionbox{\centering}{\includegraphics[width=.485\textwidth]{exp4-m1.pdf}}
\subcaptionbox{\centering 
\label{fig:e3m2}}{\includegraphics[width=.485\textwidth]{exp4-m2.pdf}}
\caption{Existence functions in Experiment 3. (\textbf{a}) submsetExistence; (\textbf{b}) supermsetExistence.\label{fig:e3m1}}
\end{figure}

@@

The data are taken from the English dictionary, which contains 235,883 different words. 
These words are mapped to multisets of integers according to the $\phi.$ In 
particular, we are interested in cases where $\phi(\Sigma)$ enumerates 
letters by their relative frequency in the English language. We say that $\phi(\Sigma)$ 
maps letters in \emph{ascending order} if the most frequent letter is mapped to 
number $\sigma.$ Conversely, in \emph{descending order}, this letter is mapped to 
the number $1.$ The size of the alphabet $\sigma$ is set to the size of the English 
alphabet: 26. The degree of a node $n$ is set to 10. On average, the multiplicity 
of letters is, of course, less than 10. We choose such a large node degree allowing 
the multiplicity to be up to 10 because the dictionary contains such words.

@@

The experiment uses the input data for the construction of the given
data structure and the test data for the execution of the operations
on the given data structure. The input data comprise a set of
randomly generated multisets used for the construction of a data
structure. The test data include the set of multisets together with
the operations that are evaluated. The input and test data were
generated with respect to parameters $\sigma$ and $n$, as presented in
%MDPI: Please confirm if this can be removed. %RES Yes, we removed it.
 Table~\ref{t:benchmark}.

\begin{table}[H]
\caption{Configuration %MDPI: We removed the vertical lines. Please confirm. %RES We confirm.
 of $\sigma$ and $n$ in benchmark.}
\label{t:benchmark}
\newcolumntype{C}{>{\centering\arraybackslash}X}
\begin{tabularx}{\textwidth}{CC}
\toprule
\boldmath$\sigma$ & \boldmath$n$ \\
\midrule
5		& 1\\
%\hline
30	& 1 \\
%\hline
5		& 3 \\
%\hline
15	& 3 \\
%\hline
30	& 3 \\
%\hline
10	& 10 \\
\bottomrule
\end{tabularx}
\end{table}
